---
title: "Teacher mental health in the 21st Century"
author: "Dr Richard W Morris"
date: "`r format(Sys.Date(), '%b %d, %Y')`"
output:
  html_document:
    theme: cosmo
    highlight: tango
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
bibliography: references.bib
csl: apa-6th-edition.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE)

library(tidyverse)
library(ggthemes)

source("~/Documents/R/helpers.R")
```


<br><br><br><br><br><br><br>

# Abstract  

The mental health and wellbeing of teachers is an important predictor of student outcomes, workforce retention as well as a broader indicator of our socioeconomic priorities. Recent changes and challenges to public schooling in Australia, including shifting community expectations and school lockdowns due to the global COVID-19 pandemic, have highlighted the connection between job strain and mental health. Here we describe that connection and how it has changed in the past 20 years in Australia using large nationally representative survey annual survey results from HILDA.  

<br><br>

# Methods  

<br>

## Variable definitions  

<br>

### Teachers (professional codes)

Teachers were defined by the ISC 4-digit codes (i.e., `jbm688`) available in the restricted dataset, based on the Australian and New Zealand Standard Classification of Occupations ([ANZSCO 2006](https://www.abs.gov.au/ausstats/abs@.nsf/0/E3031B89999B4582CA2575DF002DA702?opendocument)). The two codes for "Primary" and "Secondary" school teachers were used.  

<br>

### Remoteness  

The Accessibility/Remoteness Index of Australia (ARIA+) is provided in HILDA (`hhsra`) and based on the Australian Statistical Geography Standard Remoteness Area framework (Summerfield et al. 2019). We defined the remoteness of the region of each teacher and collapsed the three most remote categories into a single category to produce three levels of remoteness: Major Cities, Inner Regional and Other (*Outer regional, remote and very remote*).  

<br>

### Psychosocial characteristics of work  

The psychosocial characteristics of work as defined by @butterworth2010deriving, included three psychosocial job components (factors) from 12 items in all waves as well as an additional nine items available from Wave 5 (2005). We used all available items in each wave and constructed scores for each component by calculating the average of item responses (after reverse scoring the negatively worded items). Each component and the corresponding items are shown below. Item scores range from 1 to 7.    

#### Job demands & complexity

| code  | description | waves |
| ----- | ----------- | ----- |
| `jomms` | My job is more stressful than I had ever imagined | all waves |
| `jompi` | I fear that the amount of stress in my job will make me ill | all waves |
| `jomcd` | My job is complex and difficult | all waves |
| `jomns` | My job often required me to learn new skills | all waves |
| `jomus` | I use my skills in current job | all waves |
| `jomini`| My job requires me to take initiative | 5:20 |
| `jomfast` | I have to work fast in my job | 5:20 |
| `jomwi` | I have to work very intensely in my job | 5:20 |
| `jomtime` | I don’t have enough time to do everything in my job | 5:20 |

  
#### Job control  


| code  | description | waves |
| ----- | ----------- | ----- |
|`jomfd`| I have freedom to decide how I do my own work | all waves |
|`jomls`| I have a lot of say about what happens in my job | all waves |
|`jomfw`| I have freedom to decide when I do my work | all waves |
|`jomdw`| I have a lot of choice in deciding what I do at work | 5:20 |
|`jomflex`| My working times can be flexible | 5:20 |
|`jombrk`| I can decide when to take a break | 5:20 |
|`jomrpt`| My job requires me to do the same things over and over again | 5:20 |
| `jomvar`| My job provides me with a variety of interesting things to do | 5:20 |


#### Job security

| code  | description | waves |
| ----- | ----------- | ----- |
|`jompf`| I get paid fairly for the things I do in my job | all waves |
|`jomsf`| I have a secure future in my job | all waves |
|`jomcsb`| Company I work for will still be in business in 5 years | all waves |
|`jomwf`| I worry about the future of my job | all waves |

<br>

We also constructed a job resources score from the following items  

#### Job resources  

(to be done)




```{r import_data}
occupation_key = c(
  # `2310` = "College",
  `2320` = "Secondary",
  `2331` = "Primary",
  # `2332` = "Pre-primary",
  # `2340` = "Special ed",
  # `2359` = "Other teaching n.e.c."
  .default = NA_character_
)


teachers <- read_rds("../data/preprocessed.rds") %>%
  mutate(
    remoteness = fct_collapse(hhsra, 
                              `Remote` = "Outer regional", 
                              `Remote` = "Remote", 
                              `Remote` =  "Very remote"),
    remoteness = fct_recode(remoteness, `City` = "Major city", 
                                        `Regional` = "Inner regional"),
    remoteness = fct_relevel(remoteness, "City"),
    profession = recode_factor(occupation, !!!occupation_key),
    profession = fct_relevel(profession, "Primary")
  ) %>% 
  filter(!is.na(profession)) 
```

# Survey results    

##### Figure 1. Mental health has declined among teachers in city and remote regions  
```{r figure_1, fig.dim=c(9,11)}
p1 <- teachers %>%
  select(xwaveid, year, profession, remoteness, `Mental health` = ghmh, `Life satisfaction` = losat) %>%
  mutate(`Life satisfaction` = `Life satisfaction` * 10) %>%
  gather("outcome", "val", `Mental health`, `Life satisfaction`) %>%
  na.omit() %>%
  ggplot(aes(x = year, y = val, color = outcome)) +
    stat_summary(fun = "mean", geom = "line", size = 1, alpha = 0.2) +
    geom_smooth(aes(y = val), method = "loess", se=F, span = 1.5,
                formula = 'y ~ x') +
    facet_grid(remoteness~profession) +
    labs(x = "", y = "") +
    theme(legend.position = "bottom",
          legend.title = element_blank())

p1 + 
  scale_color_wsj() +
  labs(title = "Changes in mental health (blue) and life satisfaction (red) among teachers, 
across sector and region\n",
       ) +
  theme_economist() + 
  theme(legend.position = "none",
        legend.title = element_blank())
```

<br>

Average levels of mental health have declined among teachers in city and remote regions, especially recently. Life satisfaction appears to be relatively steady.  


<br><br>

##### Figure 2. Job strain has increased across sectors and regions  
```{r figure_2, fig.dim=c(9,11)}

#### Psychosocial components ####
psychosocial = c("job_control", "job_demand", "job_security")

teachers %>%
  select(xwaveid, year, remoteness, profession, all_of(psychosocial)) %>%
  gather("outcome", "val", !!!psychosocial) %>%
  filter(profession %in% c("Secondary", "Primary")) %>%
  na.omit() %>%
  ggplot(aes(x = year, y = val, color = remoteness)) +
    stat_summary(fun = "mean", geom = "line", size = 1, alpha = 0.2) +
    geom_smooth(formula = 'y ~ x', method = "loess", se=F, span=1.5) +
    facet_grid(outcome~profession, scales = "free_y") +
    labs(title = "Changes in job control, demands and security across teaching sectors 
and regions\n", 
         x = "", y = "") +
    theme_economist() +
    scale_color_wsj() +
    theme(legend.position = "bottom",
          legend.title = element_blank(),
          plot.title = element_text(vjust = 2),
          strip.text = element_text(vjust = 1.2))
```

<br>

Job strain increases with high demand and low control, and demand has been increasing while control decreasing for teachers across sectors and regions over the last 20 years. That suggests that increasing job strain is widespread.    

<br><br>

## Model results    

```{r}
library(broom)
```

Determining the role of job strain in the changes to mental health requires estimating the effect of each job characteristic on mental health in a regression model (e.g., `mental_health ~ 1 + job_control + job_demand + job_security`). Linear models of mental health were estimated for each year, and the size and the direction of the coefficients (betas) for each job characteristic indicated their importance for differences in mental health each year.  

<br><br>


##### Figure 3. Job demands (and control) are increasingly important for mental health  
```{r figure_3, fig.dim=c(9,11)}
teachers %>%
  mutate(losat = c(scale(losat)),
         ghmh = c(scale(ghmh))) %>%
  arrange(year) %>%
  group_by(year) %>%
  nest() %>%
  mutate(
    ghmh.fit = map(data, ~lm(
      formula = ghmh ~ 1 + job_control + job_demand + job_security, 
      data = .)
      ),
    losat.fit = map(data, ~lm(
      formula = losat ~ 1 + job_control + job_demand + job_security, 
      data = .)
      ),
    ghmh.betas = map(ghmh.fit, tidy),
    losat.betas = map(losat.fit, tidy),
  ) %>%
  select(year, ghmh.betas, losat.betas) -> df

bind_rows(
  select(df, year, ghmh.betas) %>%
    unnest(ghmh.betas) %>%
    mutate(model = "Mental health"),
  select(df, year, losat.betas) %>%
    unnest(losat.betas) %>%
    mutate(model = "Life satisfaction")
  ) %>%
  mutate(model = fct_relevel(model, "Mental health")) %>%
  filter(term %notin% "(Intercept)") %>%
  ggplot(aes(x = year, y = estimate, color = term, alpha = model)) +
    geom_hline(aes(yintercept = -0.1), color = "#d5e4eb", size = 0) +
    geom_hline(aes(yintercept = 0), color = "white", size = 1.25) +
    geom_hline(aes(yintercept = 0.1), color = "#d5e4eb", size = 0) +
    geom_line() +
    geom_pointrange(aes(ymin = estimate - std.error, 
                        ymax = estimate + std.error)) +
    facet_grid(term~model, scales = "free_y") +
    labs(title = "Effects (betas) of job characteristics on mental health 
and life satisfaction in each year\n", 
         x = "", y = "") +
    theme_economist() +
    scale_color_wsj() +
    theme(legend.position = "none",
          panel.grid.major.y = element_line(size = 0)) +
    scale_alpha_manual(values = c(1, 0.85))

```

<br>

Mental health seems to be increasingly dependent on job control and job demands in the last five years, as the beta estimates get further from zero between 2015 to 2020. The dependence of life satisfaction on job characteristics has remained  relatively steady over the same time period.  

<br><br>

##### Figure 4. The importance of job demands (and control) has increased in cities and regions  
```{r figure_4, fig.dim=c(9,11)}
teachers %>%
  mutate(losat = c(scale(losat)),
         ghmh = c(scale(ghmh))) %>%
  arrange(year) %>%
  group_by(year) %>%
  nest() %>%
  mutate(
    fit = map(data, ~lm(
      formula = ghmh ~ 1 + job_control:remoteness + job_demand:remoteness + 
        job_security:remoteness,
      data = .)),
    betas = map(fit, tidy)
  ) %>%
  select(year, betas) %>%
  unnest(betas) %>%
  filter(term %notin% "(Intercept)") %>%
  mutate(
    region = case_when(
      str_detect(term, "City") ~ "City",
      str_detect(term, "Regional") ~ "Regional",
      str_detect(term, "Remote") ~ "Remote"),
    job = case_when(
      str_detect(term, "job_control") ~ "job_control",
      str_detect(term, "job_demand") ~ "job_demand",
      str_detect(term, "job_security") ~ "job_security")
    ) %>%
  ggplot(aes(x = year, y = estimate, color = job)) +
    geom_hline(aes(yintercept = 0), color = "white", size = 1.25) +
    geom_line(aes(group = term)) +
    geom_pointrange(aes(ymin = estimate - std.error, 
                        ymax = estimate + std.error)) +
    facet_grid(job~region, scales = "free_y") +
    labs(title = "Effects (betas) of job characteristics on mental health 
across regions\n",
         x = "", y = "") +
    theme_economist() +
    scale_color_wsj() +
    theme(legend.position = "none",
          panel.grid.major.y = element_line(size = 0))
```

<br><br>

##### Figure 5. Job demands are more important for differences in mental health between people and over time  
```{r figure_5, fig.dim=c(9, 6)}
teachers %>%
  group_by(xwaveid) %>%
  mutate(cghmh = c(scale(ghmh, scale = F))) %>%
  ungroup() %>%
  arrange(year) %>%
  group_by(year) %>%
  nest() %>%
  mutate(
    fit = map(data, ~lm(
      formula = ghmh ~ 1 + job_control + job_demand + job_security,
      data = .)
    ),
    cfit = map(data, ~lm(
      formula = cghmh ~ 1 + job_control + job_demand + job_security,
      data = .)
    ),
    betas = map(fit, tidy),
    cbetas = map(cfit, tidy)
  ) %>%
  select(year, betas, cbetas) -> df

bind_rows(
  df %>%
    select(-cbetas) %>%
    unnest(betas) %>%
    mutate(model = "between"),
  df %>%
    select(-betas) %>%
    unnest(cbetas) %>%
    mutate(model = "within")
) %>%
  filter(term %notin% "(Intercept)") %>%
  ggplot(aes(x = year, y = estimate, color = model)) +
    geom_hline(aes(yintercept = 0), color = "white", size = 1.25) +
    geom_line(aes(group = model)) +
    geom_pointrange(aes(ymin = estimate - std.error, 
                        ymax = estimate + std.error)) +
    facet_wrap(~term) +
    labs(title = "Effects (betas) for between- (red) and within- (blue) persons\n",
           x = "", y = "") +
    theme_economist() +
    scale_color_wsj() +
    theme(legend.position = "none",
          panel.grid.major.y = element_line(size = 0))
```

<br>

Job demands and job control have become more important for explaining differences in mental health between people each year, as well as changes in mental health over time for each person.  

<br><br>

In general, the model results show that job demands and job control are increasingly impacting the mental health of teachers in cities and regions of Australia.  

<br><br>


## Tables  


##### Table 1. Change in teacher demographics between 2001 and 2020
```{r table_1}
library(gtsummary)
library(flextable)

t1 <- teachers %>% 
  filter(profession == "Primary", year %in% c(2001, 2020)) %>%
  mutate(income = hh_disp_inc_eq/1000) %>%
  mutate(Total = 1) %>%
  select(xwaveid, year, 
         Total,
         Female, 
         Age = age, 
         Coupled = coupled, 
         `New parent` = new_parent, # child under 4
         Edu = edu, 
         Region = remoteness,
         `Real household income ($000s)` = income, # OECD adjusted
         `Mental health` = ghmh,
         `Life satisfaction` = losat
  ) %>%
  distinct() %>%
  select(-xwaveid) %>%
  tbl_summary(
    by = year,
    statistic = Total ~ "N = {N}",
    type = list(`Life satisfaction` ~ "continuous"),
    missing = "no"
  ) %>%
  modify_header(all_stat_cols() ~ "**{level}**") %>%
  add_p(test.args = all_tests("fisher.test") ~ list(simulate.p.value=T),
        include = -Total)

t2 <- teachers %>% 
  filter(profession == "Secondary", year %in% c(2001, 2020)) %>%
  mutate(income = hh_disp_inc_eq/1000) %>%
  mutate(Total = 1) %>%
  select(xwaveid, year, 
         Total,
         Female, 
         Age = age, 
         Coupled = coupled, 
         `New parent` = new_parent, # child under 4
         Edu = edu, 
         Region = remoteness,
         `Real household income ($000s)` = income, # OECD adjusted
         `Mental health` = ghmh,
         `Life satisfaction` = losat
  ) %>%
  distinct() %>%
  select(-xwaveid) %>%
  tbl_summary(
    by = year,
    statistic = Total ~ "N = {N}",
    type = list(`Life satisfaction` ~ "continuous"),
    missing = "no"
  ) %>%
  modify_header(all_stat_cols() ~ "**{level}**") %>%
  add_p(test.args = all_tests("fisher.test") ~ list(simulate.p.value=T),
        include = -Total)

tbl_stack(list(t1, t2),
          group_header = c("Primary", "Secondary")) %>%
  as_flex_table() %>%
  align(j = 3:5, align = "right", part = "all") %>%
  autofit()
```


<br>

Both Primary and Secondary teachers had significant changes in education levels and income between 2001 and 2020. Secondary teachers also had significant changes in regional distribution.  

<br><br>

##### Table 2. Primary & Secondary teacher differences in 2001 & 2020 

```{r table_2}
t1 <- teachers %>%
  filter(year == 2001) %>%
  mutate(income = hh_disp_inc_eq/1000) %>%
  mutate(Total = 1) %>%
  select(xwaveid, profession, 
         Total,
         Female, 
         Age = age, 
         Coupled = coupled, 
         `New parent` = new_parent, # child under 4
         Edu = edu, 
         Region = remoteness,
         `Real household income ($000s)` = income, # OECD adjusted
         `Mental health` = ghmh,
         `Life satisfaction` = losat
  ) %>%
  distinct() %>%
  select(-xwaveid) %>%
  tbl_summary(
    by = profession,
    statistic = Total ~ "N = {N}",
    missing = "no"
  ) %>%
  modify_header(all_stat_cols() ~ "**{level}**") %>%
  add_p(test.args = all_tests("fisher.test") ~ list(simulate.p.value=T),
        include = -Total)

t2 <- teachers %>% 
  filter(year == 2020) %>%
  mutate(income = hh_disp_inc_eq/1000) %>%
  mutate(Total = 1) %>%
  select(xwaveid, profession, 
         Total,
         Female, 
         Age = age, 
         Coupled = coupled, 
         `New parent` = new_parent, # child under 4
         Edu = edu, 
         Region = remoteness,
         `Real household income ($000s)` = income, # OECD adjusted
         `Mental health` = ghmh,
         `Life satisfaction` = losat
  ) %>%
  distinct() %>%
  select(-xwaveid) %>%
  tbl_summary(
    by = profession,
    statistic = Total ~ "N = {N}",
    type = list(`Life satisfaction` ~ "continuous"),
    missing = "no"
  ) %>%
  modify_header(all_stat_cols() ~ "**{level}**") %>%
  add_p(test.args = all_tests("fisher.test") ~ list(simulate.p.value=T),
        include = -Total)

tbl_stack(list(t1, t2),
          group_header = c("2001", "2020")) %>%
  as_flex_table() %>%
  align(j = 3:5, align = "right", part = "all") %>%
  autofit() 
```

<br>

Primary school teachers tended to be more female and less educated than secondary teachers. This was consistent in 2001 and 2020.  

<br><br>

##### Table 3. Regional differences in 2020 

```{r table_3}
t1 <- teachers %>% 
  filter(profession == "Primary", year == 2020) %>%
  mutate(income = hh_disp_inc_eq/1000) %>%
  mutate(Total = 1) %>%
  select(xwaveid,  
         Total,
         Female, 
         Age = age, 
         Coupled = coupled, 
         `New parent` = new_parent, # child under 4
         Edu = edu, 
         Region = remoteness,
         `Real household income ($000s)` = income, # OECD adjusted
         `Mental health` = ghmh,
         `Life satisfaction` = losat
  ) %>%
  distinct() %>%
  select(-xwaveid) %>%
  tbl_summary(
    by = Region,
    statistic = Total ~ "N = {N}",
    type = list(`Life satisfaction` ~ "continuous"),
    missing = "no"
  ) %>%
  modify_header(all_stat_cols() ~ "**{level}**") %>%
  add_p(test.args = all_tests("fisher.test") ~ list(simulate.p.value=T),
        include = -Total)

t2 <- teachers %>% 
  filter(profession == "Secondary", year == 2020) %>%
  mutate(income = hh_disp_inc_eq/1000) %>%
  mutate(Total = 1) %>%
  select(xwaveid,  
         Total,
         Female, 
         Age = age, 
         Coupled = coupled, 
         `New parent` = new_parent, # child under 4
         Edu = edu, 
         Region = remoteness,
         `Real household income ($000s)` = income, # OECD adjusted
         `Mental health` = ghmh,
         `Life satisfaction` = losat
  ) %>%
  distinct() %>%
  select(-xwaveid) %>%
  tbl_summary(
    by = Region,
    statistic = Total ~ "N = {N}",
    type = list(`Life satisfaction` ~ "continuous"),
    missing = "no"
  ) %>%
  modify_header(all_stat_cols() ~ "**{level}**") %>%
  add_p(test.args = all_tests("fisher.test") ~ list(simulate.p.value=T),
        include = -Total)

tbl_stack(list(t1, t2),
          group_header = c("Primary", "Secondary")) %>%
  as_flex_table() %>%
  align(j = 3:5, align = "right", part = "all") %>%
  autofit() 
```


<br>

Primary school teachers in the city are morely likely to have a postgraduate degree than regional or remote primary teachers in 2020. No other differences were significant.  

<br><br>

# References  



