---
title: "The growing effect of job strain on teacher mental health"
date: "`r format(Sys.Date(), '%b %d, %Y')`"
output: word_document
bibliography: references.bib
csl: apa-6th-edition.csl
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      ft.align = "left"
                      )

library(tidyverse)
library(mgcv)
library(broom)
library(patchwork)
library(gtsummary)
library(flextable)

source("~/Documents/R/helpers.R")

gtsummary::theme_gtsummary_compact(font_size = 9)

theme_set(
  theme_minimal(base_size=18) +
  theme(legend.position = "none",
        legend.title = element_blank(),
        # panel.grid.major.y = element_line(size = 1),
        panel.grid.minor.y = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        axis.line.x = element_line(colour = "black"),
        axis.ticks.x = element_line(colour = "black"),
        axis.ticks.length.x = unit(-0.25, "cm"),
        axis.text.x = element_text(vjust = .8, hjust = 0.8, angle = 30))
)
```


Richard W. Morris^a,b^, Lisa E. Kim^c^, Alyssa Milton^a,b^, Nick Glozier^a,b^  

a.  ARC Centre of Excellence for Children and Families over the Life Course, Australia
b.  Central Clinical School, Faculty of Medicine and Health, University of Sydney, NSW, Australia
c.  Department of Education, University of York, UK


<br>

<br>

*Richard W. Morris    

**Email:**  richard.morris@sydney.edu.au  

<br><br>

**Author Contributions:** RWM designed and performed the analysis, interpreted the results, and wrote the paper; LK, AM & NG conceived the study, designed the study, interpreted the results and wrote the paper.  

**Competing Interest Statement:** No competing interests.  

**Classification:** Social Sciences/Demography.  

**Keywords:** teacher mental health, job strain, job demands, job control, HILDA.  

<br>

<br>

**This file includes:**  

- Main text, references and appendix
- Figures 1 to 4
- Tables 1  

\newpage

# Abstract  

The mental health and wellbeing of teachers is an important predictor of student outcomes, workforce retention, as well as a broader indicator of our socioeconomic priorities. However there have been reports that teachers’ mental health has been declining, particularly since the COVID-19 pandemic and shifting community expectations. The connection between employee mental health and the workplace has been established by multiple studies showing that the psychosocial work conditions are risk factors for poor mental health. Jobs with high demands and low control (i.e., autonomy) create job strain which leads to poor mental health. To explain the widely-reported decline in mental health among teachers, we determined whether the prevalence of psychosocial risk factors such as job demands have increased, or alternatively, whether the sensitivity to job demands among teachers has increased. That is, have teachers become more vulnerable due to increased prevalence or greater sensitivity to risk factors? We mapped the dependency between mental health and psychosocial risk factors for teachers and how it has changed between 2005 to 2022 using data from the Household, Income and Labour Dynamics in Australia (HILDA) survey, a large nationally representative annual survey. We find the mental health of teachers has declined since 2011 (which is not due to changes in workforce composition), along with increasing prevalence of risk factors such as job strain and greater sensitivity to job demands.  This has resulted in increased vulnerability to high job demands, which explains the decline in teacher mental health since 2011. Improving teacher resilience to high job demands is the most likely strategy to mitigate the decline in teacher mental health, however further research identifying the source(s) of the vulnerability is warranted. (271 words)  
 

\newpage

The mental health and wellbeing of teachers is an important predictor of student outcomes, workforce retention, as well as a broader indicator of our socioeconomic priorities [@mccallum2017; @braun2020; @dreer2023; @skaalvik2018; @pap2023]. Recent changes and challenges to public schooling around the world, including shifting community expectations and school lockdowns due to the global COVID-19 pandemic, have highlighted the connection between job strain and mental health in teachers (as well as in other public services such as hospitals and nurses). Since the pandemic, there have been reports that teachers mental health has been declining in the UK, USA, Canada and Australia. However whether that decline is the result of changes in the _workplace_ such as increasing job demands; or due to changes in the _workforce_ such as increased psychological vulnerability, is currently unclear. Any proposal to arrest or reverse the decline will need to address this distinction in order to be effective.    

Prior to COVID, cross-sectional comparisons of teachers with other professional groups in large population-based surveys did not reveal consistent deficits in mental health [@jerrim2020]. Since then a systematic review of studies conducted in 2020/2021 in China, Brazil, the United States, India, and Spain found anxiety, depression, and stress to be highly prevalent among teachers during the pandemic [@silva2021]. For instance, a repeated cross-sectional survey of over two million unique Facebook users between 2020 and 2021 revealed teachers in the United States were more likely to report anxiety symptoms than those in other professions [@kush2022]. Cross-sectional comparisons also revealed that anxiety and distress were higher than prepandemic levels or the national average among teachers in Canada [@hutchison2022].  

Longitudinal comparisons of mental health have also revealed declines among teachers.  Longitudinal qualitative trajectory analysis exploring teachers’ mental health at three time points in 2020 found consistent declines in teachers’ mental health in the UK [@kim2022]. Teachers self-reported mental health also declined between 2019 and 2020, before and after the outbreak of the pandemic, in a large population-based survey of Australian households [@hoang2023]. While teachers were the only distinct occupation whose mental health significantly declined over the period (relative to doctors, engineers, accountants, nurses and lawyers), such declines are also consistent with a general deterioration in mental health in the wider population over the same period [@botha2023; butterworth2022]. So despite the widely-reported decline in teacher mental health levels, it is difficult to determine whether this is due to occupation-related factors or represents a decline common to the wider population due to non-work related factors.  

The psychosocial characteristics of work are an important determinant of mental (and physical) health. Combinations of adverse job conditions are associated with poor physical health, greater depression, anxiety and more frequent visits to a general practitioner [Broom, 2006; Chandola, Brunner & Marmot 2006; Griffin et al 2007; Stansfeld & Candy, 2006); cited in @butterworth2010]. Among the most widely acknowledged risk factors are high job demands and low job control, as described by the job strain model [@karasek1998]. This model proposes that job strain results from the co-occurence of high job demands and low job control, and this in particular has been shown to be a good predictor of adverse physical and mental health problems such as coronary heart disease, depression, and anxiety.  

This report identifies whether the observed decline in teacher mental health in Australia between 2005 to 2022 is work-related, and determines whether the decline reflects an increase in the prevalence of risk factors leading to poor mental health (high job demands, low job control, or their interaction as job strain), or instead a greater sensitivity among modern teachers. I.e., Has the job got worse or have teachers become less resilient? We examine these issues among primary and secondary school teachers and contrast them to two other occupations chosen for their comparable levels of mental health (accountants) or psychosocial job characteristics (nurses & midwives).  



\newpage

# Methods  

<br>

The Household, Income and Labour Dynamics in Australia (HILDA) Survey, funded by the Australian government, was collected yearly between 2001 to 2022. The sampling frame is a nationally representative probabilistic sample of Australian households which collects information on a wide range of aspects of life in Australia, including health and wellbeing [@wilkins2019household]. The longitudinal design includes the same households and individuals each year and adds children (15-years and above) born to sample members and people who joined or shared a household with a sample member.  

<br>

## Variables  

<br>

### Occupation (professional codes)

Occupation was ascertained by the current occupation response and coded to the Australian and New Zealand Standard Classification of Occupations ([ANZSCO 2006](https://www.abs.gov.au/ausstats/abs@.nsf/0/E3031B89999B4582CA2575DF002DA702?opendocument)) 4-digit codes. The two codes for "Primary" and "Secondary" school teachers were used.  

For the sake of comparison, we also included the codes for two other well-defined professions: the code for accountants and the code for nurses & midwives. Accountants were chosen because they exhibited lower levels of mental health and life satisfaction than any other professional group in our earlier study [@hoang2023]. Nurses & midwives (hereon _nurses_) were included because they exhibited comparable levels of mental health as teachers, along with comparable levels of job demands and job control - two important prerequisites for job strain.  

<br>

### Mental health   

The 5-item mental health inventory (MHI-5) is a well-validated screening instrument for mental health in population research [@cuijpers2009; @hoeymans2004; @rumpf2001). In Australia it has been used as a proxy for common mental disorders where it has high levels of sensitivity and specificity, particularly for mood disorders [@batterham2018]. The MHI-5 is collected every year in the HILDA survey, and consists of five items assessing positive and negative aspects of mental health. Respondents are asked to state how often they have experienced each of the following during the past four weeks:  

1. "Been a nervous person"
2. "Felt so down in the dumps nothing could cheer you up"
3. "Felt calm and peaceful"
4. "Felt down"
5. "Been a happy person"  

The response to each item was selected from a 6-point scale "All of the time", "Most of the time", "A good bit of the time", "Some of the time", "A little of the time", "None of the time". The summed mental health score was created according to Ware et al [@ware2000sfu]. Each response was scored 0 to 5, and items were recoded so that higher scores indicated better mental health. Raw scores were summed across the items and then linearly transformed to a 0-100 scale. In accordance with the manual instructions a person-specific score was estimated in any year on which there were valid responses on three or more items, the average being calculated and applied to missing items.  

<br>

### Trends in mental health over time  

Trends in mental health among teachers, accountants and nurses between 2005 to 2022 were modelled as a (smooth) function of time for each occupation. The smooth trend for each occupation was estimated by a penalized, non-linear spline function over time (years) in a generalized additive model [GAM; @wood2004; @wood2006; @wood2011]. Both linear and non-linear effects of time were included and the resulting (centered) smooth trend represents the total mental health trend over time for each occupation. We also estimated an adjusted smooth trend after taking into account changes in composition of age, job tenure and gender ratio over the period as controls. Tenure was determined from the HILDA variable derived from their time with current employer cross-referenced against any occupation changes since last interview. These controls were included as linear terms.  Sensitivity analyses indicated the controls were not consequential for teacher mental health trends, but were important when estimating the mental health trends of accountants and nurses (see Figure 1).   

<br>



### Psychosocial characteristics of work  

The psychosocial characteristics of work as defined by @karasek1998 were derived from a set of 17 questions representing _job demands & complexity_ (hereon "job demands") and _job control_ [@butterworth2010]. The subset of items contributing to each component are shown in the Appendix (Tables A1 to A2). We constructed a component score for each person in each year by calculating the (within-person) average of the component items (after reverse scoring the negatively worded items). Missing responses were implicitly imputed by this mean estimation procedure. A job strain component score was calculated by the product of the other two components to represent their interaction. All 17 items were only available from Wave 5 onwards in HILDA and thus our analysis covered the period from 2005 to 2022.     

<br>

### Prevalence of risk factors (high job demand, low job control, job strain)  

The prevalence of each risk factor was determined by the proportion of people reporting high job demand, or low job control, or both (i.e., job strain) in each year. Fixed thresholds for high and low scores were defined by the response quartile of each component in the entire employed sample of HILDA (all occupations among all employed people in all seventeen waves), so as to provide a common threshold value across the three professions examined here. Thus the fixed threshold for low job control was a job control score below the first quartile of all job control responses (i.e., < 3.25), and the fixed threshold for high job demand was a job demand score above the third quartile of all job demand responses (> 4.55). Using these thresholds we defined four mutually exclusive groups of people experiencing different risk factors: 1) Following Karasek, a job strain group was defined as people reporting the co-occurrence of high job demand and low job control [@karasek1998] (i.e., job demand score > 3rd quartile of all job demand responses _and_ job control score < 1st quartile of all job control responses); 2) a high job demand only group (job demand score > 3rd quartile of all job demand responses _and_ job control score > 1st quartile of all job control responses), 3) a low job control only group (job control score < 1st quartile of all job control responses _and_ job demand score < 3rd quartile of all job demand responses); and 4) a low job strain group who reported both no high job demand and no low job control (job demand score < 3rd quartile of all job demand responses _and_ job control score > 1st quartile of all job control responses). Mutually-exclusive categories allowed us to calculate the total prevalence of risk factors in each occupation (high job demand % + low job control % + job strain %), as well as the prevalence of each risk factor separately.  

<br>

### Sensitivity to psychosocial job characteristics  

The sensitivity of teacher’s mental health to each psychosocial job characteristic was determined by estimating and comparing slopes between each job component and mental health. The slope between each component and mental health was modelled using a Generalized Additive Model (GAM), with a penalised smooth interaction term between time (years) and each job component score (see Appendix: Model definitions). This interaction term allows for a flexible, non-linear relationship between the job component and mental health, which is modulated by time (years). It will reveal whether the influence of each job component on mental health is not constant but varies with time. We standardized each job component to have a zero mean and unit variance over the seventeen years of data to allow comparison between the three smooth job component temporal interactions. The penalty ensures each smooth trend does not deviate too far from the linear slope due to variation in any particular year. We also estimated the uncertainty around each slope trend, which results in confidence intervals with close to nominal (frequentist) coverage properties [@marra2012; @nychka1988]. The greater the trend deviates from zero over time ($y = 0$), the greater the importance of that component on mental health and thus the greater vulnerability to that risk factor. Each model also included linear and smooth trends for changes in the mean level of mental health over time.  

Along with comparing the slopes for each job component, we also determined how much variance ($R^2$) in mental health was explained by all three psychosocial job components in each year. This tells us whether the importance of the job components for mental health _in toto_ is changing over time.  

<br>

### Decomposition analysis  

To decompose whether job demands, job control, or their combination in job strain, was driving the decline in teacher mental health, we compared the total mental health trend after adding each component as an explanatory variable in a model of teacher mental health. Adding each component to the model produced a residual smooth trend for mental health, which represents the mental health trend remaining after accounting for the effect of each component. We compared the residual mental health trend with the total mental health trend to determine whether any single component was responsible for, or explained, the decline in teacher mental health after accounting for changes in prevalence and sensitivity over time.

<br>

<br>

```{r import_data}
occupation_key = c(
  `2320` = "Teachers", # Secondary
  `2331` = "Teachers", # Primary
  `2230` = "Nurses",
  `2411` = "Accountants",
  .default = NA_character_
)

# Regions were collapsed according to this report:
# https://www.aihw.gov.au/reports/rural-remote-australians/rural-and-remote-health

occupations <- read_rds("../data/occupations.rds") %>% 
  filter(occupation %in% names(occupation_key)) %>%
  filter(wave %in% letters[5:22]) %>%
  mutate(
    remoteness = fct_collapse(hhsra, 
                              `Remote` = "Outer regional", 
                              `Remote` = "Remote", 
                              `Remote` =  "Very remote"),
    remoteness = fct_recode(remoteness, `City` = "Major city", 
                                        `Regional` = "Inner regional"),
    remoteness = fct_relevel(remoteness, "City"),
    profession = recode_factor(occupation, !!!occupation_key),
    profession = fct_relevel(profession, "Teachers")
  ) %>% 
  filter(!is.na(profession)) # exclude any non-teaching years
```

```{r model_funs}
# Function to extract the smooth trends
smooth_data <- function(.fit, i = 1) {
  
  plot_obj <- plot(.fit, seWithMean = T, pages=1)
  
  as.data.frame(plot_obj[[i]][c("x", "se", "fit")]) %>%
    as_tibble()
  
}

fit_ghmh <- function(.df, .prof = "Teachers", .f) {
  
  # select the data we need
  df <- filter(.df, profession == .prof) 
  
  
  # Estimate the linear and non-linear trends of year (with controls)
  fit <- gam(formula = .f, 
                data = df)
  
  return(fit)
  
}
```
\newpage

# Results  

<br>

## Survey results    

#### Table 1. Change in teacher demographics between 2005 and 2022

```{r table_1}
theme_gtsummary_compact(set_theme = TRUE, font_size = NULL)

create_table <- function(.df) {
  
  .df %>% 
    mutate(income = hh_disp_inc_eq/1000) %>%
    mutate(Total = 1) %>%
    select(xwaveid, year, 
           Total,
           Female, 
           Age = age, 
           Coupled = coupled, 
           `New parent` = new_parent, # child under 4
           Edu = edu, 
           `Tenure (years)` = jbempt,
           Region = remoteness,
           `Real household income ($000s)` = income, # OECD adjusted
           `Mental health` = ghmh,
           job_control, 
           job_demand
           ) %>%
    mutate(
      # `Low job control` = job_control < 3.25 & job_demand <= 4.556,
      # `High job demands` = job_demand > 4.556 & job_control >= 3.25,
      `Job strain` = job_control < 3.25 & job_demand > 4.556
    ) %>%
    distinct() %>%
    select(-xwaveid, -job_demand, -job_control) %>%
    tbl_summary(
      by = year,
      statistic = Total ~ "N = {N}",
      missing = "no"
    ) %>%
    modify_header(all_stat_cols() ~ "**{level}**") %>%
    add_p(test.args = all_tests("fisher.test") ~ list(simulate.p.value=T),
          include = -Total)
  
}

tbl <- occupations %>%
  filter(year %in% c(2005, 2022)) %>%
  group_by(profession) %>%
  nest() %>%
  mutate(tb = map(data, ~create_table(.x)))

tbl_stack(tbl$tb,
          group_header = tbl$profession) %>%
  as_flex_table() %>%
  align(j = 3:5, align = "right", part = "all") %>%
  hrule(rule = "exact", part = "all") %>%
  height_all(height = 0.2) %>% 
  width(j = 1, width = 0.9) %>%
  width(j = 2, width = 2.2) %>%
  width(j = 3:4, width = 1.2) %>%
  width(j = 5, width = 0.75)
```


<br>

Teachers had significant changes in mean age, education levels, and income between 2005 and 2022. The increasing prevalence of degree requirements across the States and Territories increased the proportion of teachers with a bachelors degree by 2022, and this has been reflected by increased compensation (real household income). Nurses experienced similar levels and changes in education and income between 2005 and 2022 as teachers. Accountants had higher median levels of education and income than teachers (and nurses), but experienced less change in median income.   

All three occupations experienced similar levels as well as declines in mental health between 2005 and 2022, however only the decline for teachers and nurses was significant. The prevalence of job strain (a major risk factor for poor mental health) increased between 2005 and 2022 for teachers and nurses, while remained relatively low and steady for accountants. We examine the changes in mental health levels and prevalence of risk factors in more detail below.  

<br>

<br>

## Trends in mental health    

We examine in greater detail the decline in mental health of teachers, nurses and accountants observed between 2005 and 2022. We estimated the total smooth trend in mental health over the period, as well as an adjusted trend after taking into account changes in composition of age, job tenure and gender ratio over the period as controls. These controls were included because we know that mental health scores of the general population change with age and gender in this data [@botha2023], and job tenure is also a likely predictor of mental health changes.   

##### Figure 1. Mean (±SEM), estimated (95%CI), and adjusted (95%CI) mental health trends from 2005 to 2022
```{r ghmh_trends_prep, include=F}
# estimate fits with and without controls
wc <- as.formula(ghmh ~ s(year) + year + age + age:year + jbempt + jbempt:year +
                  Female + Female:year)
nc <- as.formula(ghmh ~ s(year))

fit_t_nc <- fit_ghmh(occupations, "Teachers",    nc)
fit_t_wc <- fit_ghmh(occupations, "Teachers",    wc)
fit_n_nc <- fit_ghmh(occupations, "Nurses",      nc)
fit_n_wc <- fit_ghmh(occupations, "Nurses",      wc)
fit_a_nc <- fit_ghmh(occupations, "Accountants", nc)
fit_a_wc <- fit_ghmh(occupations, "Accountants", wc)

smooths <- bind_rows(
  `Teachers_estimated` = smooth_data(fit_t_nc),
  `Teachers_adjusted` = smooth_data(fit_t_wc),
  `Nurses_estimated` = smooth_data(fit_n_nc),
  `Nurses_adjusted` = smooth_data(fit_n_wc),
  `Accountants_estimated` = smooth_data(fit_a_nc),
  `Accountants_adjusted` = smooth_data(fit_a_wc),
  .id = "model"
) %>%
  separate(model, into = c("profession", "ctl"), sep = "_") %>%
  mutate(profession = fct_rev(profession))

# summarise mean and se
mean_se <- function(x, mult = 1) {
  
  x = na.omit(x)
  m = mean(x)
  se = sd(x) / sqrt(length(x))
  
  out <- data.frame(
    fit = m,
    se = se*mult
  )
  
  return(out)
}

obs <- occupations %>%
  select(xwaveid, x = year, profession, ghmh) %>%
  group_by(profession, x) %>%
  summarize(mse = mean_se(ghmh)) %>%
  unnest(mse) %>%
  mutate(ctl = "observed")
```
```{r ghmh_trends, fig.dim = c(9, 9)}
# add to plot data
to_plot <- bind_rows(smooths, obs) %>%
  mutate(ctl = fct_rev(ctl))

to_plot1 <- filter(to_plot, ctl != "observed")
to_plot2 <- filter(to_plot, ctl == "observed")

ggplot(to_plot1, aes(x = x - 2000, y = fit, ymin = fit - se, ymax = fit + se)) +
  geom_hline(aes(yintercept = 0), color = "grey50", linetype = "dashed") +
  geom_ribbon(alpha = 0.2) +
  geom_line() +
  geom_pointrange(data = to_plot2, aes(x = x - 2000, y = fit)) +
  scale_x_continuous(breaks = c(1, 5, 10, 15, 20)) +
  facet_grid(ctl~profession, scales = "free_y") +
  labs(x = "", y = "") +
  theme(legend.position = "none",
        axis.text.x = element_text(vjust = -1, hjust = 0.5, angle = 0))
```

<br>

Figure 1 top row shows the observed mean levels of mental health (MHi-5 units) of teachers, nurses and accountants declined over the period, however teachers showed a rapid decline after peaking in 2011. The middle and bottom rows of Figure 1 show the centred smooth trends for mental health. The estimated smooth mental health trend of teachers exhibits the same sudden decline after 2011 while the other two professions exhibit a more linear decline over the period. However the bottom row of Figure 1 shows that once the mental health trends are adjusted for changes in age, gender composition and job tenure, only the mental health trend of teachers significantly declined over the period.   

Tests of the non-linear changes revealed the estimated trend of teachers declined non-linearly ($F_{3.78} = 12.21$, $p < .001$), even after adjustment ($F_{3.88} = 6.97$, $p < .001$). There was no evidence of non-linear changes in the estimated trend of nurses and accountants after adjustment ($F_{1} = 0.19$, $p = .66$, and $F_{2.63} = 0.98$, $p = .53$, respectively), and a follow-up linear mixed model with random intercepts and effects of time for each participant confirmed the (fixed) linear effect of time was also non-significant for the $\beta_{time}$ of nurses as well as accountants ($t_{479.8} = -0.86$, $p = .39$ and $t_{411.5} = 0.78$, $p = .43$, respectively).  

Consequently, we focus the remainder of the paper on distinguishing the role of changes in prevalence and sensitivity to the different psychosocial risk factors on the decline in mental health of teachers.  

<br>

<br>

## Prevalence of psychosocial risk factors (high job demand, low job control, job strain)  

##### Figure 2. The cumulative proportion of people reporting each risk factor in each year
```{r psc_prevalence, fig.dim=c(9,6)}
ctrl_t <- 3.25 # with(read_rds("../data/occupations.rds"), summary(job_control))
dmnd_t <- 4.556 # with(read_rds("../data/occupations.rds"), summary(job_demand))

df <- occupations %>%
  # select(xwaveid, year, ghmh, job_control, job_demand, job_security) %>%
  mutate(
    job_strain = case_when(
      job_control < ctrl_t & job_demand > dmnd_t ~ "Job strain",
      job_control < ctrl_t ~ "Low control",
      job_demand > dmnd_t ~ "High demand",
      TRUE ~ "Low strain"),
    job_strain = fct_relevel(job_strain, "Job strain")
  )

prop.ci <-  function(n, p, a=0.05) {
  ## Level (1-a) confidence interval for proportion (equivalent to prop.test)
  z <- qnorm(1-a/2,lower.tail=FALSE)
  l <- 1/(1+1/n*z^2)*(p + 1/2/n*z^2 + 
                        z*sqrt(1/n*p*(1-p) + 1/4/n^2*z^2))
  u <- 1/(1+1/n*z^2)*(p + 1/2/n*z^2 -
                        z*sqrt(1/n*p*(1-p) + 1/4/n^2*z^2))
  c(lower=l, upper=u)
}

df %>%
  percent(profession, year, job_strain) %>%
  mutate(N = sum(n)) %>%
  filter(job_strain != "Low strain") %>%
  rowwise() %>%
  mutate(lo = prop.ci(N, proportion)[1],
         hi = prop.ci(N, proportion)[2]) %>% 
  ggplot(aes(x = year, y = proportion, group = job_strain)) +
    geom_pointrange(aes(ymin = lo, ymax = hi)) +
    geom_smooth(method = "lm", formula = 'y ~ x', 
                # adjust level for prediction intervals
                se = T, level = .9999999, size = .5, color = "black") +
    facet_grid(job_strain~profession) +
    labs(x = "", y = "")
```

<br>

Among teachers, the total prevalence of psychosocial risk factors has increased from 62 percent in 2005 to 72 percent in 2022 (+10%). Nurses have experienced a steeper increase in total prevalence from 50 percent in 2005 to 71 percent in 2022 (+21%), while prevalence among accountants has held relatively low and steady (36 to 37%). Taking each risk factor separately, the prevalence of job strain among teachers was less than twenty percent (18%) in 2005 and increased to almost 30 percent (29%) in 2022 (+11%). By comparison, the prevalence of job strain among nurses has increased from 13 to 20 percent (+7%), while accountants have held steady at a relatively low two percent. The prevalence of high job demands among teachers and accountants has been relatively steady (~25% and ~30%, respectively), while it has increased for nurses from 24 to 35 percent (+10%) between 2005 to 2022. The prevalence of low job control among teachers was 13 percent in 2005 before rising recently to 22 percent in 2021 & 2022 (+9%), while among nurses and accountants the prevalence of low job control has been relatively steady (~13% and ~3%, respectively).           

<br>

<br>


## Sensitivity to psychosocial job characteristics  

Understanding the drivers in mental health vulnerability requires we examine how sensitivity to the various job components has changed over time, in addition to any changes in prevalence. The sensitivity (or dependency) of teachers' mental health on each psychosocial job characteristic is represented by the slope of each component as a predictor of mental health (in a regression model). We present the changes over time in the slope of each component to the mental health of teachers in Figure 3 below. Each smooth trend represents the penalized effect of the psychosocial job characteristic on mental health between 2005 to 2022. Superimposed on the penalized smooth effects is the unpenalized linear effect (i.e., beta) of each component estimated by OLS for each year, as points and vertical ranges ($\beta±SEM$). This allows us to compare the penalized esimates to the unpenalized (and therefore overfit) estimates.  

<br>

##### Figure 3. Standardized effects (smooths and betas) of each component on mental health over time  

```{r smooth_psc_prep, include=F}
#### Smooth fits ####
df <- occupations %>%
  filter(profession == "Teachers") %>%
  select(xwaveid, year, ghmh, job_control, job_demand, job_security) %>%
  mutate(year = year - 2000,
         ghmh = c(scale(ghmh)),
         job_strain = job_control * job_demand) %>%
  mutate(across(job_control:job_strain, ~c(scale(.x))))
         
fit_year_psc_int <- gam(ghmh ~ 1 + job_control + job_demand + job_strain + 
                          s(year) + s(year, by = job_demand) + 
                          s(year, by = job_control) + 
                          s(year, by = job_strain), 
                        data = df)

job_smooths <- bind_rows(
  `Job demand` = smooth_data(fit_year_psc_int, 2),
  `Job control` = smooth_data(fit_year_psc_int, 3),
  `Job strain` = smooth_data(fit_year_psc_int, 4),
  .id = "term"
) %>%
  mutate(term = fct_relevel(term, "Job demand"))

#### Linear fits ####
lm_fits <- df %>%
  group_by(year) %>%
  nest() %>%
  mutate(
    ghmh.fit = map(data, ~lm(
      formula = ghmh ~ 1 + job_control + job_demand + job_strain, 
      data = .)
    ))

lm_betas <- lm_fits %>%
  transmute(
    betas = map(ghmh.fit, tidy)) %>%
  unnest(betas)
```

```{r smooth_psc, fig.dim = c(9, 6)}
job_betas <- lm_betas %>%
  filter(term %notin% c("(Intercept)", "age", "jbempt")) %>%
  mutate(term = str_replace(term, "_", " "),
         term = str_to_sentence(term)) %>%
  select(term, x = year, se = std.error, fit = estimate) %>%
  mutate(term = fct_relevel(term, "Job demand")) %>%
  ungroup()

# Fix the stupid results so control is positive and strain is negative
fix_key <- c(
  `Job control` = "Job strain",
  `Job strain` = "Job control"
)

job_betas <- job_betas %>%
  mutate(term = recode_factor(term, !!!fix_key))

job_smooths <- job_smooths %>%
  mutate(term = recode_factor(term, !!!fix_key)) 

ggplot(job_smooths, aes(x = x, y = fit)) +
  geom_hline(aes(yintercept = 0), color = "red", linetype = 2) +
  geom_ribbon(aes(ymin = fit - se, ymax = fit + se),
              alpha = 0.2) +
  geom_pointrange(data = job_betas, 
                  aes(ymin = fit - se, ymax = fit + se)) +
  geom_line() +
  scale_x_continuous(breaks = c(1, 5, 10, 15, 20)) +
  facet_wrap(~term, ncol = 3, scales = "free_y") +
  labs(x = "", y = "") +
  theme(legend.position = "none",
        axis.text.x = element_text(vjust = -1, hjust = 0.5, angle = 0))
```


<br>

Figure 3 shows the unique importance of each psychosocial job characteristic for teacher mental health over time. Importance is indicated by the deviation from $y = 0$ in each smooth trend. The location of the smooths (beta) with respect to $y = 0$ indicates the direction of the effect. Job strain and job demand had a negative effect on mental health while job control had a positive effect. Where each trend deviates away from zero with time (and where the shaded confidence intervals exclude zero) indicates where teachers become more sensitive to that component over time. For example, job demand became more important for teachers’ mental health after 2011, while the importance of job control and job strain was relatively stable over the same period.  

The OLS fit ($R^2$) of all three psychosocial job characteristics on mental health each year from 2005 to 2022 improved from ~4 percent at the beginning to almost ~16 percent by the end of the period (see Figure A1 in Appendix). The improvement in the $R^2$ suggests that psychosocial job components explain an increasing proportion of the variation in the mental health of teachers in the past seventeen years, consistent with increased vulnerability to psychosocial risk factors.  

<br>

<br>


## Decomposition results  

The relative importance of job demands for the mental health of teachers increased faster than any other psychosocial job characteristic. At the same time, the prevalence of the risk factor (high job demand) was relatively stable among teachers over the same period (while the prevalence of the other risk factors increased). This makes it difficult to distinguish whether it is the increasing prevalence of any risk factor (e.g., job control or job strain), or the greater sensitivity to any risk factor (e.g., job demand) is driving the decline in teacher mental health. To decompose whether job demands, job control, or job strain was driving the decline in mental health, we compared the total mental health trend after adding each component as an explanatory variable in a model of teacher mental health. The changes to the residual mental health trend after adding each component, relative to the total trend, reveal whether any single component was mediating the decline in mental health.  

Figure 4 below shows that teachers’ total mental health trend showed a rapid decline over time, except after the effect of job demands was explained (i.e., added to the model as a predictor).  

<br>

##### Figure 4. Total (left) and residual trends in mental health over time after adding different psychosocial job components
```{r smooth_est_prep, include=F}
# Plot to compare how the year smooth changes with the addition of the job 
# variables (similar to a mediation analysis)

df <- occupations %>%
  mutate(job_strain = job_demand * job_control)

year_f <- formula(ghmh ~ s(year) + year + 
                     age + age:year + jbempt + jbempt:year)

demand_f <- formula(ghmh ~ s(year) + year + 
                      s(year, by = job_demand) + job_demand +
                       age + age:year + jbempt + jbempt:year)

control_f <- formula(ghmh ~ s(year) + year +
                       s(year, by = job_control) + job_control +
                       age + age:year + jbempt + jbempt:year)

security_f <- formula(ghmh ~ s(year) + year +
                        s(year, by = job_security) + job_security +
                       age + age:year + jbempt + jbempt:year)

strain_f <- formula(ghmh ~ s(year) + year +
                       s(year, by = job_strain) + job_strain +
                       age + age:year + jbempt + jbempt:year)

fit_t_year <- fit_ghmh(df, .f = year_f)
fit_t_demand <- fit_ghmh(df, .f = demand_f)
fit_t_control <- fit_ghmh(df, .f = control_f)
# fit_t_security <- fit_ghmh(df, .f = security_f)
fit_t_strain <- fit_ghmh(df, .f = strain_f)


smooths <- bind_rows(
  `f(year)` = smooth_data(fit_t_year),
  `f(year) + f(demand)` = smooth_data(fit_t_demand),
  `f(year) + f(control)` = smooth_data(fit_t_control),
  `f(year) + f(strain)` = smooth_data(fit_t_strain),
  .id = "model"
) %>% 
  mutate(model = fct_relevel(model, "f(year)", "f(year) + f(strain)"))
```

```{r smooth_est, fig.dim = c(9, 6)}
ggplot(smooths, aes(x = x - 2000, y = fit)) +
  geom_hline(aes(yintercept = 0), color = "grey50", linetype = "dashed") +
  geom_ribbon(aes(ymin = fit - se, ymax = fit + se),
              alpha = 0.2) +
  geom_line(aes()) +
  scale_x_continuous(breaks = c(1, 5, 10, 15, 20)) +
  facet_wrap(~model, nrow = 1) +
  labs(x = "", y = "") +
  theme(legend.position = "none",
        axis.text.x = element_text(vjust = -1, hjust = 0.5, angle = 0))
```

<br>

Figure 4 legend. Estimated smooth mental health trends of teachers between 2005 to 2022 under different explanatory models.  

The left panel of Figure 4 ($f(year)$) shows the total smooth trend in teachers' mental health, unadjusted for any psychosocial job characteristics, and previously shown in Figure 1 which declined sharply after 2011. The right panel ($f(year) + f(demand)$) shows that once the impact of job demands is explained, the negative decline in mental health is completely reversed and mental health was predicted to rise over the period. The middle panels ($f(year) + f(strain)$ and $f(year) + f(control)$) shows the impact of job strain and job control on the decline is minimal at best, as the mental health trend declines at a similar rate with or without these components in the model. Overall this result represents model-derived evidence the decline in teachers' mental health is (entirely?) explained by changes in the effect of job demands on teachers, even after accounting for any changes in the prevalence or sensitivity to this risk factor over time.    

Testing the non-linear changes revealed there was no evidence the increase in mental health after including job demand was non-linear ($F_{3.29} = 1.65$, $p = .16$), however the linear change was positive and significant ($t_{415} = 67.47$, $p < .001$). By comparison the non-linear and linear changes in mental health remained significant after job control and job security were included (all $ps < .001$).  

<br>

<br>

# Discussion  

_Brief discussion points for consideration_  

Main point: The mental health of teachers has rapidly declined since 2011 and this seems to be explained by increasing senstivity to job demands in the last ten years rather than a change in prevalence of this risk factor.  

- The decline in mental health of teachers is not due to changes in composition of the workforce (unlike nurses and accountants)
- The prevalence of high job demands has not substantially changed, however the importance of job demands for the mental health of teachers has increased rapidly since 2011
- Residual analysis shows the negative trend in mental health reverses when the effect of job demands are explained
- The psychosocial job components may explain almost as much as 20 percent of variation ($R^2$) in teachers mental health today 

In general, we find converging evidence that increasing sensitivity to job demands, rather than increasing prevalence of high job demands, has driven poorer mental health among teachers in the last decade.   

The effects on teacher mental health are widespread, appearing in both primary and secondary teachers, and in cities and remote schools (less apparent in regional schools). 
_Implications_. As much as 20 percent of the variation in teachers' mental health may be addressed by efforts to improve teacher resilience to high job demands. This is because it was the changes in teacher sensitivity to job demands rather than changes in prevalence of high job demands which occurred alongside the declines in mental health. By contrast, reducing the prevalence of high job demands, or improving job control, is unlikely to be an effective strategy for treating vulnerability.  

_Weaknesses_. Despite the longitudinal data and our (make-shift) decomposition analysis, our methods are ultimately based on observational data and so the results are correlational and conclusions are descriptive. However it is obviously impossible to examine our research question with an experimental design, and so while policy development must depend on studies such as ours, it should be developed with caution. HILDA also misses an important moderator of job strain which is _job resources_, such as support from colleagues/management and other psychosocial support, which may govern whether interventions are successful but we cannot draw any conclusions here.  

_Strengths_. Many surveys of occupational wellbeing, or items focused on work-related wellbeing tend to produce biased estimates due to self-selection of employees unhappy with their job and seeking to express their concerns or grievances in a single-issue survey response. Even large supposedly ‘representative’ samples can over-represent unhappy respondents by more than 2.5 times due to such self-selection incentives into work-related surveys [@chauvenet2020; @goodwin2013; @harvey2021]. We used HILDA which does not suffer this problem due its wide topic coverage and probabilistic sampling.  

HILDA also incorporates a comprehensive set of questions about other aspects of subjective wellbeing, including life satisfaction. In contrast to mental health, teachers generally report higher levels of life satisfaction and self-worth than other occupations [@hoang2023; @jerrim2020]. Both mental health and life satisfaction are important but distinct components of subjective wellbeing, and the comprehensive instruments in HILDA allowed us to distinguish mental health from these other more cognitive/evaluative dimensions which can depend on social level or stage or life [@diener2017; @diener2018; @kettlewell2020]. Many conflicting findings in the literature may be a result of comparing or confounding these different constructs.    

_Importance_. Skaalvik and Skaalvik (2018) argue that teacher subjective wellbeing, which may be considered a type of job resource, predicts higher job engagement and satisfaction and lower motivation to leave the profession [@skaalvik2018]. Multiple studies have found positive relationships between teacher subjective wellbeing and student achievement, albeit mostly correlational [@mccallum2017; @braun2020; @dreer2023]; while others have linked teacher subjective wellbeing with student physical and mental wellbeing [@braun2020; @pap2023], thus expanding the role of teachers beyond merely developing knowledge to contributing to the development of student social and emotional wellbeing, including mental health [@jennings2009].  









\newpage

# Appendix  

<br>

## Variable definitions  

<br>

#### Table A1. Job demands & complexity items 

```{r}
readr::read_delim("
Code|Description|Waves
jomms|My job is more stressful than I had ever imagined|1 to 22
jompi|I fear that the amount of stress in my job will make me ill|1 to 22
jomcd|My job is complex and difficult|1 to 22
jomns|My job often required me to learn new skills|1 to 22
jomus|I use my skills in current job|1 to 22
jomini|My job requires me to take initiative|5 to 22
jomfast|I have to work fast in my job|5 to 22
jomwi|I have to work very intensely in my job|5 to 22
jomtime|I don’t have enough time to do everything in my job|5 to 22", 
delim = "|") %>%
  flextable() %>%
  fontsize(size = 9, part = "all") %>%
  flextable::theme_alafoli() %>%
  hrule(rule = "exact", part = "all") %>%
  height_all(height = 0.2) %>% 
  width(j = 1, width = 0.63) %>%
  width(j = 2, width = 3.64) %>%
  width(j = 3, width = 0.6) 
```

<br>

<br>
  
#### Table A2. Job control items    

```{r}
readr::read_delim(
"Code|Description|Waves
jomfd|I have freedom to decide how I do my own work|1 to 22
jomls|I have a lot of say about what happens in my job|1 to 22
jomfw|I have freedom to decide when I do my work|1 to 22
jomdw|I have a lot of choice in deciding what I do at work|5 to 22
jomflex|My working times can be flexible|5 to 22
jombrk|I can decide when to take a break|5 to 22
jomrpt|My job requires me to do the same things over and over again|5 to 22
jomvar`|My job provides me with a variety of interesting things to do|5 to 22",
delim = "|") %>%
  flextable() %>%
  fontsize(size = 9, part = "all") %>%
  flextable::theme_alafoli() %>%
  hrule(rule = "exact", part = "all") %>%
  height_all(height = 0.2) %>% 
  width(j = 1, width = 0.63) %>%
  width(j = 2, width = 3.64) %>%
  width(j = 3, width = 0.6) 
```

<br>

<br>

<br>

<br>

## Model definitions  

For the sensitivity to risk-factors analysis, models regressing mental health (MHi-5 scores) in each occupation onto each psychosocial job component (job demand, job control, job strain) were estimated, using non-linear, smooth interaction terms representing the temporal trend of each component in a generalized additive model (GAM):  

$$
\text{MHi-5}_{it} \sim \beta_0 + f(\text{year}_t) + f(\text{year}_t) \cdot \text{job control}_{it} + \epsilon
$$  

$$
\text{MHi-5}_{it} \sim \beta_0 + f(\text{year}_t) + f(\text{year}_t) \cdot \text{job demand}_{it} + \epsilon
$$  

$$
\text{MHi-5}_{it} \sim \beta_0 + f(\text{year}_t) + f(\text{year}_t) \cdot \text{job strain}_{it} + \epsilon
$$  


Where $\text{MHi-5}_{it}$ represents the MHi-5 score for each person in each year, $f(\text{year}_t)$ is a smooth term to capture the yearly changes in the average MHi-5 level using cubic splines, and $f(\text{year}_t) \cdot \text{job component}_{it}$ is another non-linear term representing smooth changes in the slope between the standardized job component scores and MHi-5 scores in each year over time.  




<br>

<br>

##### Figure A1. Proportion of variance in mental health explained by all risk factors (r-squared)
```{r r_squared, fig.dim=c(9, 6)}
lm_fits %>%
  transmute(
    term = "r-squared",
    r.squared = map_dbl(ghmh.fit, ~glance(.) %>% pull(r.squared))) %>%
  ggplot(aes(x = year, y = r.squared)) +
    geom_line() +
    geom_point() +
    scale_x_continuous(breaks = c(1, 5, 10, 15, 20)) +
    labs(y = "", x = "") +
    theme(legend.position = "none",
        axis.text.x = element_text(vjust = -1, hjust = 0.5, angle = 0))
```

<br>

<br>


## State-specific effects   

In Australia, state and territory governments are responsible for running schools and setting local policy while the federal government provides the national strategy. For this reason, we considered whether including state effects in the trend models was warranted. State effects were included as separate penalized trends and the variation of each trend was estimated to determine whether it was significantly different from the overall trend. This is similar to deciding whether to include random effects in a mixed model by examining the variation around the random effect. The estimated smooth trends in mental health and the psychosocial job characteristics of teachers for each state and territory are shown below.  

<br>

##### Figure A2. Estimated State-specific trends in teacher mental health
```{r state_specific_ghmh, fig.dim = c(9, 9)}
to_fit <- occupations %>%
  filter(profession == "Teachers") %>% # Teachers Nurses Accountants
  select(xwaveid, year, hhstate, ghmh) %>%
  mutate(
    ohhstate = ordered(hhstate, 
      levels = c("NSW", "VIC", "QLD", "WA", "SA", "TAS", "NT", "ACT")),
    hhstate = factor(hhstate)
    ) %>%
  na.omit() 

states <- gam(ghmh ~ 1 + hhstate + s(year, fx=T) + 
                s(year, by = hhstate), 
              method = "REML",
              data = to_fit)

no_states <- gam(ghmh ~ 1 + hhstate + s(year, fx=T),
                 method = "REML",
                 data = to_fit)

# Positive difference is advantage for no_states 
AIC_dif = round(glance(states)$AIC - glance(no_states)$AIC, 1)
BIC_dif = round(glance(states)$BIC - glance(no_states)$BIC, 1)
  
  
ggplot(to_fit, aes(x = year, y = ghmh)) +
  stat_smooth(method = "gam", size = .3, color = "black") +
  scale_x_continuous(breaks = c(2001, 2005, 2010, 2015, 2020)) +
  facet_wrap(~ohhstate, nrow = 2) +
  labs(x = "", y = "") +
  scale_color_grey(start = 0, end = .8) +
  theme(legend.position = "none",
        axis.text.x = element_text(vjust = -1, hjust = 0.5, angle = 0))
```

<br>

The state model estimated no positive mental health trends among the states, and a declining trend for six states, except for Tasmania (TAS) and the Australian Capital Territory (ACT), which represented less than 5 percent of our sample of teachers (*n* = 54). Table A6 shows the results of a likelihood ratio test against a null distribution of zero variance for each state, which reveals there was little evidence of any state or territory specific trends (smallest _p_-value = .061 for WA).  

<br>

#### Table A6. Likelihood Ratio Test results for state-specific trends in teacher mental health ($H_0 = 0$ variance)
```{r state_specific_pvalues}
states <- gam(ghmh ~ 1 + hhstate + s(year, fx=T) + 
                s(year, by = hhstate), 
              data = to_fit)

tidy(states) %>%
  filter(term %notin% "s(year)") %>%
  mutate(term = str_remove_all(term, c("s\\(year\\)\\:hhstate"))) %>%
  rename(`State/territory` = term, df = ref.df) %>%
  flextable() %>%
    colformat_double(digits = 2) %>%
    colformat_double(j = "p.value", digits = 3) %>%
    add_footer_lines("*edf: estimated degrees of freedom") %>%
    fontsize(size = 9, part = "all") %>%
    theme_alafoli() %>%
    hrule(rule = "exact", part = "all") %>%
    height_all(height = 0.2) %>%
    autofit()
```

<br>

The results of a model comparison between a State-specific model and a non-specific model confirmed there was not sufficient evidence to warrant including State-specfic effects as an explanation [@shmueli2010]. The BIC difference between the two models was `r BIC_dif`, which is a large difference favouring the non-specific model [@raftery1995] and consistent with the non-specific model. The AIC difference between the two models was `r AIC_dif`, which is a small difference and equivalent to a non-significant difference between models [@raftery1995].  

<br>

<br>

## Cumulative prevalence

#### Figure A4. Cumulative prevalence of psychosocial risk factors (job strain, high job demand, low job control)  

```{r cumulative_prevalence, fig.dim=c(9,6)}
ctrl_t <- 3.25 # with(read_rds("../data/occupations.rds"), summary(job_control))
dmnd_t <- 4.556 # with(read_rds("../data/occupations.rds"), summary(job_demand))

df <- occupations %>%
  # select(xwaveid, year, ghmh, job_control, job_demand, job_security) %>%
  mutate(
    job_strain = case_when(
      job_control < ctrl_t & job_demand > dmnd_t ~ "Job strain",
      job_control < ctrl_t ~ "Low control",
      job_demand > dmnd_t ~ "High demand",
      TRUE ~ "Low strain"),
    job_strain = fct_relevel(job_strain, "Job strain")
  )

# Cumulative proportion of professionals who meet criteria for job strain in
# each year
df %>%
  percent(profession, year, job_strain) %>%
  mutate(job_strain = fct_rev(job_strain)) %>% 
  mutate(job_strain = recode(job_strain, `Low strain` = " ")) %>% 
  ggplot(aes(x = year, y = proportion, fill = job_strain)) +
    geom_area(alpha = 0.9, size = .5, color = "white") +
    scale_x_continuous(expand = c(0.01, 0.1), 
                       breaks = c(2001, 2005, 2010, 2015, 2020)) +
    scale_y_continuous(expand = c(0.01, 0.01),
                       breaks = c(0.25, 0.50, 0.75, 1.0)) +
    scale_fill_grey(start = 1, end = 0.2) + #
    facet_wrap(~profession) +
    theme_void() +
    theme(legend.title = element_blank(),
          legend.position = "bottom",
          axis.text = element_text(color = "black"),
          strip.text = element_text(size = 12))



```

<br>

<br>

## Job security  

Others have identified job insecurity as an important risk factor and linked with psychological disstress (De Witte & Naswall 2003), however it was not prevalent in our teacher sample (see Appendix) and so not featured in this report.  

#### Figure A3. Prevalence of psychosocial risk factors (high job demand, low job control, job insecurity)  

```{r alt_prevalence, fig.dim=c(9,6), eval=F}
ctrl_t <- 3.25 # with(read_rds("../data/occupations.rds"), summary(job_control))
dmnd_t <- 4.556 # with(read_rds("../data/occupations.rds"), summary(job_demand))
scty_t <- 4.5

df <- occupations %>%
  # select(xwaveid, year, ghmh, job_control, job_demand, job_security) %>%
  mutate(
    job_strain = case_when(
      # job_control < ctrl_t & job_demand > dmnd_t ~ "Job strain",
      job_control < ctrl_t ~ "Low control",
      job_demand > dmnd_t ~ "High demand",
      job_security < scty_t ~ "Low security",
      TRUE ~ "Low strain")
    # job_strain = fct_relevel(job_strain, "Job strain")
  )

# Proportion of professionals who meet criteria for job strain in each year
df %>%
  percent(profession, year, job_strain) %>%
  mutate(job_strain = fct_rev(job_strain)) %>% 
  mutate(job_strain = recode(job_strain, `Low strain` = " ")) %>% 
  ggplot(aes(x = year, y = proportion, fill = job_strain)) +
  geom_area(alpha = 0.9, size = .5, color = "white") +
  scale_x_continuous(expand = c(0.01, 0.1), 
                     breaks = c(2001, 2005, 2010, 2015, 2020)) +
  scale_y_continuous(expand = c(0.01, 0.01),
                     breaks = c(0.25, 0.50, 0.75, 1.0)) +
  scale_fill_grey(start = 1, end = 0.2) + #
  facet_wrap(~profession) +
  theme_void() +
  theme(legend.title = element_blank(),
        legend.position = "bottom",
        axis.text = element_text(color = "black"))
```

<br>


# Not run  

<br>

#### Table A3. Job security items

```{r eval=F}
readr::read_delim(
"Code|Description|Waves
jompf|I get paid fairly for the things I do in my job|1 to 22
jomsf|I have a secure future in my job|1 to 22
jomcsb|Company I work for will still be in business in 5 years|1 to 22
jomwf|I worry about the future of my job|1 to 22",
delim = "|") %>%
  flextable() %>%
  fontsize(size = 9, part = "all") %>%
  flextable::theme_alafoli() %>%
  hrule(rule = "exact", part = "all") %>%
  height_all(height = 0.2) %>% 
  width(j = 1, width = 0.63) %>%
  width(j = 2, width = 3.64) %>%
  width(j = 3, width = 0.6) 
```

<br>

<br>

#### Table A4. Primary & Secondary teacher differences in 2005 & 2022
```{r eval=F}
t1 <- teachers %>%
  filter(year == 2005) %>%
  mutate(income = hh_disp_inc_eq/1000) %>%
  mutate(Total = 1) %>%
  select(xwaveid, profession, 
         Total,
         Female, 
         Age = age, 
         Coupled = coupled, 
         `New parent` = new_parent, # child under 4
         Edu = edu, 
         `Tenure (years)` = jbempt,
         Region = remoteness,
         `Real household income ($000s)` = income, # OECD adjusted
         `Mental health` = ghmh,
         `Life satisfaction` = losat
  ) %>%
  distinct() %>%
  select(-xwaveid) %>%
  tbl_summary(
    by = profession,
    statistic = Total ~ "N = {N}",
    type = list(`Life satisfaction` ~ "continuous"),
    missing = "no"
  ) %>%
  modify_header(all_stat_cols() ~ "**{level}**") %>%
  add_p(test.args = all_tests("fisher.test") ~ list(simulate.p.value=T),
        include = -Total)

t2 <- teachers %>% 
  filter(year == 2022) %>%
  mutate(income = hh_disp_inc_eq/1000) %>%
  mutate(Total = 1) %>%
  select(xwaveid, profession, 
         Total,
         Female, 
         Age = age, 
         Coupled = coupled, 
         `New parent` = new_parent, # child under 4
         Edu = edu, 
         `Tenure (years)` = jbempt,
         Region = remoteness,
         `Real household income ($000s)` = income, # OECD adjusted
         `Mental health` = ghmh,
         `Life satisfaction` = losat
  ) %>%
  distinct() %>%
  select(-xwaveid) %>%
  tbl_summary(
    by = profession,
    statistic = Total ~ "N = {N}",
    type = list(`Life satisfaction` ~ "continuous"),
    missing = "no"
  ) %>%
  modify_header(all_stat_cols() ~ "**{level}**") %>%
  add_p(test.args = all_tests("fisher.test") ~ list(simulate.p.value=T),
        include = -Total)

tbl_stack(list(t1, t2),
          group_header = c("2005", "2022")) %>%
  as_flex_table() %>%
  fontsize(size = 9, part = "all") %>%
  align(j = 3:5, align = "right", part = "all") %>%
  hrule(rule = "exact", part = "all") %>%
  height_all(height = 0.2) %>% 
  width(j = 1, width = 0.9) %>%
  width(j = 2, width = 2.2) %>%
  width(j = 3:4, width = 1.2) %>%
  width(j = 5, width = 0.75)
```

<br>

Primary school teachers tended to be more female, younger, and less postgraduate qualifications than secondary teachers. This was consistent across both 2005 and 2022.  

<br>

<br>


#### Table A5. Regional differences in 2022
```{r eval=F}
t1 <- teachers %>% 
  filter(profession == "Primary", year == 2022) %>%
  mutate(income = hh_disp_inc_eq/1000) %>%
  mutate(Total = 1) %>%
  select(xwaveid,  
         Total,
         Female, 
         Age = age, 
         Coupled = coupled, 
         `New parent` = new_parent, # child under 4
         Edu = edu, 
         `Tenure (years)` = jbempt,
         Region = remoteness,
         `Real household income ($000s)` = income, # OECD adjusted
         `Mental health` = ghmh,
         `Life satisfaction` = losat
  ) %>%
  distinct() %>%
  select(-xwaveid) %>%
  tbl_summary(
    by = Region,
    statistic = Total ~ "N = {N}",
    type = list(`Life satisfaction` ~ "continuous"),
    missing = "no"
  ) %>%
  modify_header(all_stat_cols() ~ "**{level}**") %>%
  add_p(test.args = all_tests("fisher.test") ~ list(simulate.p.value=T),
        include = -Total)

t2 <- teachers %>% 
  filter(profession == "Secondary", year == 2022) %>%
  mutate(income = hh_disp_inc_eq/1000) %>%
  mutate(Total = 1) %>%
  select(xwaveid,  
         Total,
         Female, 
         Age = age, 
         Coupled = coupled, 
         `New parent` = new_parent, # child under 4
         Edu = edu, 
         `Tenure (years)` = jbempt,
         Region = remoteness,
         `Real household income ($000s)` = income, # OECD adjusted
         `Mental health` = ghmh,
         `Life satisfaction` = losat
  ) %>%
  distinct() %>%
  select(-xwaveid) %>%
  tbl_summary(
    by = Region,
    statistic = Total ~ "N = {N}",
    type = list(`Life satisfaction` ~ "continuous"),
    missing = "no"
  ) %>%
  modify_header(all_stat_cols() ~ "**{level}**") %>%
  add_p(test.args = all_tests("fisher.test") ~ list(simulate.p.value=T),
        include = -Total)

tbl_stack(list(t1, t2),
          group_header = c("Primary", "Secondary")) %>%
  as_flex_table() %>%
  fontsize(size = 9, part = "all") %>%
  align(j = 3:5, align = "right", part = "all") %>%
  hrule(rule = "exact", part = "all") %>%
  height_all(height = 0.2) %>% 
  width(j = 1, width = 0.75) %>%
  width(j = 2, width = 1.9) %>%
  width(j = 3:5, width = 1.2) %>%
  width(j = 6, width = 0.75)
```


<br>

Primary school teachers in the city were more likely to have a postgraduate degree than regional or remote primary teachers in 2022. No other differences were significant. 

```{r not_run_0, eval=F, fig.dim=c(9,6)}
# Figure 2. Mean (±SEM) levels of the psychosocial job components from 2005 to 
# 2022

psychosocial = c("job_control", "job_demand", "job_security")

occupations %>%
  filter(profession == "Teachers") %>%
  select(xwaveid, year, all_of(psychosocial)) %>%
  gather("outcome", "val", !!!psychosocial) %>%
  mutate(outcome = str_replace(outcome, "_", " "),
         outcome = str_to_title(outcome),
         outcome = fct_relevel(outcome, "Job Demand")) %>%
  na.omit() %>%
  ggplot(aes(x = year, y = val)) +
    stat_summary(fun.data = ggplot2::mean_se) +
    stat_smooth(method = "gam", #se=F, span=1.5,
                geom = "line", size = .75) +
    scale_x_continuous(breaks = c(2001, 2005, 2010, 2015, 2020)) +
    facet_wrap(~outcome, scales = "free_y") +
    labs(x = "", y = "") +
    theme(legend.position = "bottom")

# Figure 2 shows that teachers' job demands have increased over time while job 
# control has decreased, however the mean levels of job security have varied 
# around 5.5. Job strain occurs with low job control and high job demands, 
# suggesting that job strain may have been increasing over time among teachers, 
# which we examine further below.   
```



```{r not_run_1, eval=F}
to_fit <- occupations %>%
  filter(profession == "Teachers") %>% 
  # select ghmh job_control job_demand job_security for y below:
  select(xwaveid, year, hhstate, y = job_security) %>% 
  mutate(
    ohhstate = ordered(hhstate, 
      levels = c("NSW", "VIC", "QLD", "WA", "SA", "TAS", "NT", "ACT")),
    hhstate = factor(hhstate)
    ) %>%
  na.omit() 

state_specific <- gam(y ~ 1 + hhstate + s(year) + 
             s(year, by = hhstate), 
           data = to_fit,
           method = "REML")

tidy(state_specific)


nonspecific <- gam(y ~ 1 + hhstate + s(year),
           data = to_fit,
           method = "REML")

BIC(state_specific, nonspecific) # smaller BIC is better model

psychosocial = c("job_control", "job_demand", "job_security")

occupations %>%
  filter(profession == "Teachers") %>% # Teachers Nurses Accountants
  select(xwaveid, year, hhstate, all_of(psychosocial)) %>%
  gather("outcome", "val", !!!psychosocial) %>%
  mutate(outcome = str_replace(outcome, "_", " "),
         outcome = str_to_title(outcome),
         outcome = fct_relevel(outcome, "Job Demand"),
         hhstate = fct_relevel(hhstate, "NSW", "VIC", "QLD", "WA", "SA")) %>%
  na.omit() %>%
  ggplot(aes(x = year, y = val, color = outcome)) +
    stat_summary(fun.data = mean_se) +
    stat_smooth(formula = 'y ~ x', method = "loess", se=F, span=1.5,
                geom = "line", size = .75) +
    scale_x_continuous(breaks = c(2001, 2005, 2010, 2015, 2020)) +
    facet_wrap(~hhstate, nrow = 2, scales = "free_y") +
    labs(x = "", y = "") +
    scale_color_grey(start = 0, end = .8) +
    theme(legend.position = "bottom")

# We also performed a similar analysis for state-specific effects in trends in 
# each of the psychosocial job components. There was marginal evidence of a 
# state-specific trend in job control in VIC (p = .05), however no other
# state-specific trends were significantly different from zero variance and the
# BIC difference was 58.82 favouring a non-specific model. There was no evidence
# for state-specific trends in job demands (smallest p-value = .401 for NSW, BIC
# difference was 57.96 favoring a non-specific model) nor job security (smallest
# p-value = .06 for TAS, BIC difference was 89.28 favouring a non-specific 
# model)
```


```{r not_run_2, eval=F, fig.dim=c(9,11)}
# Mental health (dark) but not life satisfaction (light) has declined among
# teachers in city and remote regions  

to_plot <- teachers %>%
  select(xwaveid, year, profession, remoteness, `Mental health` = ghmh, `Life satisfaction` = losat) %>%
  mutate(`Life satisfaction` = `Life satisfaction` * 10) %>%
  gather("outcome", "val", `Mental health`, `Life satisfaction`) %>%
  na.omit() 

p1 <- to_plot %>%
  ggplot(aes(x = year, y = val, color = outcome)) +
    stat_summary(fun.data = mean_se) +
    geom_smooth(aes(y = val), method = "loess", se=F, span = 1.5, size = .5,
                formula = 'y ~ x') +
    facet_grid(remoteness~profession) +
    labs(x = "", y = "") +
    theme(legend.position = "bottom",
          legend.title = element_blank()) +
  scale_color_grey(start = 0.8, end = 0.2) +
  theme(legend.position = "none")

p1
```


```{r not_run_3, eval=F, fig.dim=c(9,11)}
# Job strain from demands and control has increased across sectors and regions  

#### Psychosocial job components ####
psychosocial = c("job_control", "job_demand", "job_security")

teachers %>%
  select(xwaveid, year, remoteness, profession, all_of(psychosocial)) %>%
  gather("outcome", "val", !!!psychosocial) %>%
  mutate(outcome = str_replace(outcome, "_", " "),
         outcome = str_to_title(outcome)) %>%
  na.omit() %>%
  ggplot(aes(x = year, y = val, color = remoteness)) +
    stat_summary(fun.data = mean_se) +
    stat_smooth(formula = 'y ~ x', method = "loess", se=F, span=1.5, 
                geom = "line", alpha = 0.7, size = 1) +
    facet_grid(outcome~profession, scales = "free_y") +
    labs(x = "", y = "") +
    scale_color_grey(start = 0, end = .8) +
    theme(legend.position = "bottom")
```


```{r not_run_4, eval=F, fig.dim = c(9, 6)}
# Mental health levels (Mean ±SEM) among teachers by job strain levels over time
ggplot(df, aes(x = year - 2000, y = ghmh)) +
  stat_summary(aes(color = job_strain), fun.data = mean_se) +
  geom_smooth(aes(color = job_strain), se=F) +
  scale_x_continuous(breaks = c(1, 5, 10, 15, 20)) +
  # stat_summary(aes(color = job_strain), geom = "line", fun = mean) +
  # stat_summary(aes(color = job_strain), geom = "point", fun = mean,
  #              shape = 21, size = 3, fill = "white", stroke = 1) +
  scale_color_grey() +
  facet_wrap(~job_strain, nrow = 1) +
  labs(x = "", y = "") 

# Teachers defined as suffering job strain in each year tend to report lower 
# mental health levels than other teachers, however the effect of high job 
# demands and low job control on mental health appears to be additive and does
# not interact.   
```




\newpage

# References  



