---
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      fig.align = "center"
                      )

library(tidyverse)
library(mgcv)
library(broom)
library(patchwork)

source("~/Documents/R/helpers.R")

gtsummary::theme_gtsummary_compact(font_size = 9)

theme_set(
  theme_minimal(base_size=18) +
  theme(legend.position = "none",
        legend.title = element_blank(),
        # panel.grid.major.y = element_line(size = 1),
        panel.grid.minor.y = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        axis.line.x = element_line(colour = "black"),
        axis.ticks.x = element_line(colour = "black"),
        axis.ticks.length.x = unit(-0.25, "cm"),
        axis.text.x = element_text(vjust = -1, hjust = 0.5, angle = 0),
        axis.title.x = element_text(size = 14, vjust = -1),
        axis.title.y = element_text(size = 14)
        # axis.text.x = element_text(vjust = .8, hjust = 0.8, angle = 30)
        )
)
```

```{r import_data}
occupation_key = c(
  `2320` = "Teachers", # Secondary
  `2331` = "Teachers", # Primary
  `2230` = "Nurses",
  # `2411` = "Accountants",
  .default = NA_character_
)

# Regions were collapsed according to this report:
# https://www.aihw.gov.au/reports/rural-remote-australians/rural-and-remote-health

occupations <- read_rds("../data/teachers & nurses.rds") %>% 
  filter(occupation %in% names(occupation_key)) %>%
  filter(wave %in% letters[5:22]) %>%
  mutate(
    remoteness = fct_collapse(hhsra, 
                              `Remote` = "Outer regional", 
                              `Remote` = "Remote", 
                              `Remote` =  "Very remote"),
    remoteness = fct_recode(remoteness, `City` = "Major city", 
                                        `Regional` = "Inner regional"),
    remoteness = fct_relevel(remoteness, "City"),
    profession = recode_factor(occupation, !!!occupation_key),
    profession = fct_relevel(profession, "Teachers"),
    sex = as.factor(if_else(Female, "female", "male"))
  ) %>% 
  filter(!is.na(profession)) # exclude any non-teaching years
```

```{r fit_ghmh}
fit_ghmh <- function(.df, .prof = "Teachers", .f) {
  
  # select the data we need
  df <- filter(.df, profession == .prof) 
  
  
  # Estimate the linear and non-linear trends of year (with controls)
  fit <- gam(formula = .f, 
                data = df)
  
  return(fit)
  
}
```

```{r smooth_data}
# Function to extract the smooth trends
smooth_data <- function(.fit, i = 1) {
  
  # seWithMean = T: The component smooths are shown with confidence intervals 
  # that include the uncertainty about the overall mean as well as the centred
  # smooth itself (as suggested by Marra & Wood 2012) 
  plot_obj <- plot(.fit, seWithMean = T, pages=1)
  
  as.data.frame(plot_obj[[i]][c("x", "se", "fit")]) %>%
    as_tibble()
  
}
```

### Figure 1. Yearly mental health and prevalence of psychosocial risk factors from 2005 to 2022
```{r plot_prevalence, fig.dim = c(9, 9)}
ctrl_t <- 3.25 # with(read_rds("../data/occupations.rds"), summary(job_control))
dmnd_t <- 4.556 # with(read_rds("../data/occupations.rds"), summary(job_demand))


# summarise mean and ci
percent_ci <- function(.x) {
  # must pass a numeric vector!
  
  .x <- stats::na.omit(.x)
  x = sum(.x)
  n = sum(!is.na(.x))
  
  prop.test(x = x, n = n) %>%
    broom::tidy() %>%
    select(y = estimate, ymin = conf.low, ymax = conf.high) %>%
    mutate(across(everything(), ~. *100)) %>%
    mutate(n = n)
}

bind_rows(
  `MHI-5 score (0-100)` = occupations %>%
    group_by(profession, year) %>%
    summarise(
      result = mean_se(ghmh, mult = 1.96),
      .groups = "drop"
    ) %>%
    unnest(result),
  
  `low control (%)` = occupations %>%
    mutate(low_control = job_control < ctrl_t) %>%
    group_by(profession, year) %>%
    summarise(
      result = percent_ci(low_control),
      .groups = "drop"
    ) %>%
    unnest(result),
  
  `high demand (%)` = occupations %>%
    mutate(high_demand = job_demand > dmnd_t) %>%
    group_by(profession, year) %>%
    summarise(
      result = percent_ci(high_demand),
      .groups = "drop"
    ) %>%
    unnest(result),
  
  .id = "outcome"
) %>%
  mutate(outcome = fct_relevel(outcome, 
          "MHI-5 score (0-100)", "low control (%)")) -> fig1


ggplot(fig1, aes(x = year, y = y)) +
  geom_pointrange(aes(ymin = ymin, ymax = ymax)) +
  # geom_smooth(se=F, linetype = "dashed", linewidth = .5, colour = "black") +
  facet_grid(outcome ~ profession, scales = "free_y") +
  labs(y = "", x = "
       Survey year") +
  scale_y_continuous(labels = scales::number_format(accuracy = 1))
```

<br>

#### Figure 1 legend.  
Mean MHI-5 scores (0-100) and prevalence (%) of people reporting _low_ job control and _high_ job demand in each year (where _high_ and _low_ are defined by the job component score thresholds between the most extreme quartile of the employed population). Vertical lines represent 95 percent confidence intervals.  


\newpage

### Figure 2. Mental health trend analysis partial effects 
```{r Figure_2, include=F}
model.df <- occupations %>%
  mutate(rjob_control = 7 - job_control, # reversed (0 = complete job control)
         ajob_demand = job_demand - 1,   # adjusted (0 = no job demands)
         cage = c(scale(age, scale=F)),
         cjbempt = c(scale(jbempt, scale=F)))

# Model definitions ####
# Using a continuous by variable allows the smooth for year to interact with the
# parametric job term. Such models are sometimes called varying coefficient 
# models. The idea is that the linear regression coefficient for year is varying
# smoothly with job (or vice versa).
# 
# Partial effects control the included variables at zero, so including reversed
# job control or reversed job demands estimates the effect on mental health when
# job control is 
# 
# (Must include jbempt, uncentered so partial effects work below)
unadj_f <- formula(ghmh ~ s(year))
adjst_f <- formula(ghmh ~ s(year) + 
                     cage + cage:year + sex + sex:year +
                     jbempt + jbempt:year)
cntrl_f <- formula(ghmh ~ s(year) + 
                     cage + cage:year + sex + sex:year +
                     jbempt + jbempt:year +
                     s(year, by = rjob_control) + rjob_control)
demnd_f <- formula(ghmh ~ s(year) + cage + cage:year + sex + sex:year +
                     jbempt + jbempt:year +
                     s(year, by = ajob_demand) + ajob_demand)


fit_t_unadj <- fit_ghmh(model.df, .prof = "Teachers", .f = unadj_f)
fit_t_adjst <- fit_ghmh(model.df, .prof = "Teachers", .f = adjst_f)
fit_t_cntrl <- fit_ghmh(model.df, .prof = "Teachers", .f = cntrl_f)
fit_t_demnd <- fit_ghmh(model.df, .prof = "Teachers", .f = demnd_f)
fit_n_unadj <- fit_ghmh(model.df, .prof = "Nurses",   .f = unadj_f)
fit_n_adjst <- fit_ghmh(model.df, .prof = "Nurses",   .f = adjst_f)
fit_n_cntrl <- fit_ghmh(model.df, .prof = "Nurses",   .f = cntrl_f)
fit_n_demnd <- fit_ghmh(model.df, .prof = "Nurses",   .f = demnd_f)

# Partial effect plots ####
# Plot the zero-centred, partial effect of year (i.e., the partial-effect sets
# all other terms to zero)
# 
# https://ecogambler.netlify.app/blog/interpreting-gams/
#
# main plot
df_t <- bind_rows(
  `Model I` = smooth_data(fit_t_unadj),
  `Model II` = smooth_data(fit_t_adjst),
  `Model III` = smooth_data(fit_t_cntrl),
  `Model IV` = smooth_data(fit_t_demnd),
  .id = "model"
) %>%
  mutate(profession = "Teachers")

df_n <- bind_rows(
  `Model I` = smooth_data(fit_n_unadj),
  `Model II` = smooth_data(fit_n_adjst),
  `Model III` = smooth_data(fit_n_cntrl),
  `Model IV` = smooth_data(fit_n_demnd),
  .id = "model"
) %>%
  mutate(profession = "Nurses")
```
```{r plot_trend_analysis, fig.dim=c(9,6)}
fig2 <- bind_rows(df_t, df_n) %>%
  mutate(profession = fct_relevel(profession, "Teachers"))

ggplot(fig2, aes(x = x, y = fit)) +
  geom_hline(aes(yintercept = 0), color = "grey50", linetype = "dashed") +
  geom_ribbon(aes(ymin = fit - se, ymax = fit + se),
              alpha = 0.2) +
  geom_line(aes()) +
  # facet_grid(model ~ profession) +
  facet_grid(profession ~ model) +
  labs(y = "Centered MHI-5 trend (±95%CI)", x = "
       Survey year") +
  theme(axis.title = element_text(size = 14))
```

<br>

#### Figure 2 legend.  
Zero-centered partial effects of year (±95%CI) on mental health from each model. Model I is the unadjusted trend over time. Model II is the trend adjusted for non-work related factors (age, sex, length of job tenure). Model III is adjusted for the same non-work related factors as well as job control. Model IV is adjusted for job demands (and the non-work related factors). Shaded areas represent 95 percent confidence intervals.   



\newpage

### Figure 3. Sensitivity of teacher mental health to job demands and job control  
```{r ghmh_sensitivity, include=F}
model.df <- occupations %>%
  filter(profession == "Teachers") %>%
  transmute(
    xwaveid,
    year = year - 2000,
    ghmh,
    job_control,                  # 0 = no job control
    rjob_demand = 8 - job_demand, # reverse job demands so effects are positive
    sex,
    cage = c(scale(age, scale=F)),
    cjbempt = c(scale(jbempt, scale=F)))
  
# Both stressors are included in the same model .: relative importance
fit_year_psc_int <- gam(ghmh ~ 1 + job_control + rjob_demand +  
                          s(year) + s(year, by = rjob_demand) + 
                          s(year, by = job_control),
                        data = model.df)

job_smooths <- bind_rows(
  `Job demand` = smooth_data(fit_year_psc_int, 2),
  `Job control` = smooth_data(fit_year_psc_int, 3),
  .id = "term"
) 

lm_fits <- model.df %>%
  group_by(year) %>%
  nest() %>%
  mutate(
    ghmh.fit = map(data, ~lm(
      formula = ghmh ~ 1 + job_control + rjob_demand, 
      data = .)
    ))

lm_betas <- lm_fits %>%
  transmute(
    betas = map(ghmh.fit, tidy)) %>%
  unnest(betas)

job_betas <- lm_betas %>%
  filter(term %notin% c("(Intercept)", "age", "jbempt")) %>%
  mutate(
    # fix the coefficient names
    term = str_replace(term, "_", " "),
    term = str_remove(term, "^r"),
    term = str_to_sentence(term),
    # center the OLS estimates
    estimate = if_else(term == "Job demand", estimate - 1.4, estimate - 0.6),
    std.error = std.error * 1.96
  ) %>%
  select(term, x = year, se = std.error, fit = estimate) %>%
  ungroup()
```
```{r plot_sensitivity, fig.dim=c(9,6)}
ggplot(job_smooths, aes(x = x + 2000, y = fit)) +
  geom_pointrange(data = job_betas, aes(ymin = fit - se, ymax = fit + se),
                  color = "grey30") +
  geom_ribbon(aes(ymin = fit - se, ymax = fit + se),
              alpha = 0.2) +
  geom_line() +
  facet_wrap(~term) +
  labs(x = "
       Survey year", y = "Absolute MHI-5 units (±95%CI)") 
```

<br>

#### Figure 3 legend.  
Penalized (smooth shaded region) and unpenalized (points and vertical lines) zero-centered partial-effects (±95%CI) of each job component on teacher mental health, representing sensitivity (in absolute MHI-5 units) of mental health to psychosocial stressors between 2005 and 2022. Where each 95% confidence interval excludes zero (shaded or error-bar) indicates where teachers became more sensitive to that component over time or in that year.  